---
title: "Ebola variant emergence surveillance analytics"
author: "FIRM, MH-SPIB"
date: "`r Sys.Date()`"
output: html_document
params:
  metadata: "trimmed_metadata.tsv"
  div_tree_raw: "ebola-bdbv.treefile"
  div_tree_clean: "genetic_distance_tree.nwk"
  time_tree: "temporal_tree.nwk"
  rtt_script: "rtt_fit_plot.R"
  name_col: "accessionVersion"
  date_col: "sampleCollectionDate"
  y_cut_val: 0.00081
  n_sd: 3
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(treeio)
library(ggtree)
library(dplyr)
library(stringr)
library(ggplot2)
library(readr)
```

```{r unpack-params, echo=FALSE}
# unpack params into standalone variables used throughout this report --------
metadata_path  <- params$metadata
div_tree_raw   <- params$div_tree_raw
div_tree_clean <- params$div_tree_clean
time_tree_path <- params$time_tree
rtt_script     <- params$rtt_script
name_col       <- params$name_col
date_col       <- params$date_col
y_cut_val      <- params$y_cut_val
n_sd           <- params$n_sd
```

```{r load-meta, echo=FALSE}
# load and clean metadata ------------------------------------------------
meta <- read.delim2(metadata_path)
meta[[date_col]] <- as.Date(meta[[date_col]])
# note: sequences known to be problematic (poor quality, ADAR editing, etc.)
# are expected to have already been removed upstream.

# make a coarse Location variable: keep country for non-DRC, otherwise
# collapse DRC admin-2 areas into a few groups of interest.
bunia_rwampara <- c("bunia", "rwampara")
mongbwalu      <- c("mongbwalu", "mongwalu", "mungwalu")

df <- meta
if (all(c("geoLocAdmin2", "geoLocCountry") %in% colnames(df))) {
  df <- df %>%
    mutate(
      admin2_lower  = str_to_lower(str_trim(geoLocAdmin2)),
      country_lower = str_to_lower(str_trim(geoLocCountry)),
      Location = case_when(
        str_detect(country_lower, "democratic republic|congo|drc") &
          admin2_lower %in% bunia_rwampara ~ "Bunia_Rwampara",
        str_detect(country_lower, "democratic republic|congo|drc") &
          admin2_lower %in% mongbwalu      ~ "Mongbwalu",
        str_detect(country_lower, "democratic republic|congo|drc") ~ "rest_DRC",
        TRUE ~ str_trim(geoLocCountry)
      )
    ) %>%
    select(-admin2_lower, -country_lower)
} else {
  df$Location <- NA_character_
}
```

```{r load-trees, echo=FALSE}
# diversity tree (iqtree result, no clock outliers removed)
div_tree1        <- read.tree(div_tree_raw)
tree_path_divraw <- div_tree_raw

# diversity tree after clock outliers removed (treetime genetic distance tree)
div_tree2          <- read.tree(div_tree_clean)
tree_path_divclean <- div_tree_clean

# time tree (treetime temporal tree)
time_tree <- read.tree(time_tree_path)

# load the root-to-tip regression function
source(rtt_script)

# manual thresholds (dataset dependent)
# (y_cut_val already unpacked above)
```

## Checking for successful clades

Looking for any long branches or any dominating clades.

Plot of the diversity tree (full data, no clock outliers removed):

```{r trees-div, fig.height = 12, fig.width = 8, echo=FALSE}
p <- ggtree(div_tree1) %<+% df +
  geom_tippoint(aes(colour = Location), size = 2, alpha = 1) +
  geom_tiplab(size = 0) +
  scale_color_brewer(palette = "Set1") +
  theme_tree2()
p
```

(number of sequences in tree = `r length(div_tree1$tip.label)`)

The time tree after clock outliers are removed:

```{r trees-time, fig.height = 12, fig.width = 8, echo=FALSE}
p3 <- ggtree(time_tree, mrsd = max(meta[[date_col]], na.rm = TRUE)) %<+% df +
  geom_tippoint(aes(colour = Location), size = 2, alpha = 1) +
  geom_tiplab(size = 0) +
  scale_color_brewer(palette = "Set1") +
  theme_tree2()
p3
```

(number of sequences in tree = `r length(time_tree$tip.label)`)

The number of sequences removed by the iterative TreeTime run is
`r length(div_tree1$tip.label) - length(time_tree$tip.label)`.

## Testing for evolutionary jumps

1. Using the diversity tree, plot root-to-tip with linear regression (`r n_sd` SD shaded region).
2. Find tips at the top right of the plot, outside the SD envelope (collect names).
3. Re-plot the diversity and time tree with the outliers highlighted. If they
   cluster in a clade coming off a long-ish branch, it could be an emerging variant.

### 1. Find clock outliers

Plot with the entire data set (both simple and best-fit regression), no outliers removed:

```{r r2t-raw, echo=FALSE}
# simple regression (given root)
result1a <- root_to_tip_regression(
  tree_path = tree_path_divraw,
  metadata  = meta,
  name_col  = name_col,
  date_col  = date_col,
  best_root = FALSE,
  n_sd      = n_sd
)
result1a$summary
result1a$outliers
result1a$plot

# best fit (search for root maximising R^2)
result1b <- root_to_tip_regression(
  tree_path = tree_path_divraw,
  metadata  = meta,
  name_col  = name_col,
  date_col  = date_col,
  best_root = TRUE,
  n_sd      = n_sd
)
result1b$summary
result1b$outliers
result1b$plot

outliers1b <- result1b$outliers
```

### 2. Examine tips in the top-right corner of the plot

Gold points are the most diverse sequences sitting above the expected clock rate.

```{r top-quad, echo=FALSE}
x_cut <- median(result1b$data$date_num)
y_cut <- y_cut_val

df1 <- result1b$data
top_right <- df1 %>%
  filter(date_num > x_cut, rtt > y_cut)

target_points <- df1[df1$tip %in% top_right$tip, ]

result1b$plot +
  geom_point(
    data        = target_points,
    aes(x = date_num, y = rtt),
    color       = "gold",
    size        = 2.2,
    alpha       = 0.9,
    inherit.aes = FALSE
  )
```

Are they clustering above the expected clock rate? If yes, this could be an
evolutionary jump.

### 3. Re-plot the trees with these target points identified

Are these same sequences clustering in a single clade? If yes, it might be an
emerging variant; if no, then not.

In the time tree:

```{r timetree-targets, fig.height = 12, fig.width = 8, echo=FALSE}
treeplotdata <- p3$data
target_tips  <- treeplotdata[treeplotdata$label %in% target_points$tip, ]

p3 +
  geom_tippoint(data = target_tips, color = "black", size = 4.5) +
  geom_tippoint(data = target_tips, color = "gold",  size = 4)
```

In the diversity tree:

```{r divtree-targets, fig.height = 12, fig.width = 8, echo=FALSE}
treeplotdata1 <- p$data
target_tips1  <- treeplotdata1[treeplotdata1$label %in% target_points$tip, ]

p +
  geom_tippoint(data = target_tips1, color = "black", size = 4.5) +
  geom_tippoint(data = target_tips1, color = "gold",  size = 4)
```

```{r cluster-info, echo=FALSE}
# metadata for the identified candidate cluster
meta %>%
  filter(.data[[name_col]] %in% target_points$tip) -> cluster_info
cluster_info
```

Interpretation notes: if the candidate sequences span multiple locations and a
wide date range they are less likely to represent a single local outbreak.
Recent collection dates within a tight cluster increase the likelihood of a
mutational advantage worth monitoring. Continue surveillance for additional
recent cases/sequences in any flagged cluster.

## Methods notes

Using Ebolaflow output. Data and metadata from Pathoplexus. Root-to-tip fitting
based on Clockor2 methods (https://github.com/clockor2/clockor2).